Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-24 Manuscript/manuscript.Rmd

Cell type-specific genomic annotations were used to build the predictor space for of our random forest models. We created three types of predictors: *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--Additional performance metrics including AUROC, accuracy, sensitivity, and specificity were [??? are?] reported in the supplementary materials.-->

@@

@@

Expectedly, when using data as-is, that is, with class imbalance present, the models performed poorly. Using any of the resampling techniques was found to improve model performance, with RUS narrowly out performing SMOTE re-sampling as the most optimal class balancing technique for all resolutions (Figure 3). These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. <!--These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.-->

Additionally, we found that regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 2). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

Interestingly, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be due to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Additionally, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. In summary, our results imply that using random under- and over-sampling, distance-type predictors, and the data at 10kb resolution results in an optimally performing cell line-specific random forest-based TAD boundary prediction model.

@@

## Transcription factor binding sites outperform other annotation-specific models

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three groups of genomic annotations (histone modifications (HM), chromatin states (ChromHMM), and transcription factor binding sites (TFBS)). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 3). Furthermore, models build solely on TFBSs outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 4A). Notably, the TFBS-specific models outperformed models built on the combined sets of annotations, and this effect was most pronounced at lower resolutions. This indicated that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

<!--
![**Assessing top ranked transcription factors in TAD boundary prediction.** (A) Barplots comparing performances of models using specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three groups (Combined; purple). (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes. Error bars represent standard deviation from the mean MCC across each chromosome.](Figures/annot_specific_rfe_varimps_heatmap.png)
-->

@@

>>>>

Cell type-specific genomic annotations [???"annotations" or "elements"] were used to build the predictor space for of our random forest models. We created four types of predictors: *Peak Signal Strength* - [???include definition] (refered to hereafter as *Signal*), *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. [???include reason for using mcc, why not others?] <!--Additional performance metrics including AUROC, accuracy, sensitivity, and specificity were reported in the supplementary materials.-->

@@

\newpage

@@

Expectedly, when using data as-is, that is, with class imbalance present, the models exhibited low MCCs. Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types (Figure 3). These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. <!--These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.-->Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types (Figure 3). As with comparing class balancing techniques, this improvement was less evident at lower resolutions. <!--This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.-->

Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.

@@

## Transcription factor binding sites outperform HM- and CRE-specific models

We then aimed to compare the performances of genomic class-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three classes of genomic elements including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). For an optimally built random forest classifier at each data resolution--smote re-sampling, and distance-type predictors--we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites for both cell lines (Supplementary Figure 3 A-B). Furthermore, models built solely on TFBS outperformed other class-specific models (Figure 4A; Supplementary Figure 4A). Notably, the TFBS-specific models outperformed models built on the combination of all genomic elements at lower resolutions, indicating that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

![**Assessing top ranked transcription factors in TAD boundary prediction.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (CRE; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three classes (Combined; purple). (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. Error bars represent standard deviation from the mean MCC across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes.](Figures/annots_specifc_rfe_importances.png)

@@

<!--Additional performance metrics were reported for both cell lines including positive predictive value (PPV; precision), F1-Score (F1), Accuracy, Area Under the Receiver Operating Characteristic Curve (AUC), Area Under the Precision-Recall Curve.-->
